---
title: "Adaptation of AD QSP model to include PopPK implementation for STEP Tutorial"
author: "Blerta Shtylla and Luke Fostvedt"
date: "09/18/2026"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE, fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
library(tibble)
library(purrr)
```

Source: [nlmixr2lib vignette](https://nlmixr2.github.io/nlmixr2lib/articles/Miyano_2022_atopicDermatitis_qsp.html) ([Rmd source](https://github.com/nlmixr2/nlmixr2lib/blob/main/vignettes/articles/Miyano_2022_atopicDermatitis_qsp.Rmd))

## Model and source

- Citation: Miyano T, Irvine AD, Tanaka RJ. A mathematical model to identify optimal combinations of drug targets for dupilumab poor responders in atopic dermatitis. Allergy. 2022;77(2):582-594.
- Article: https://doi.org/10.1111/all.14870
- PubMed: https://pubmed.ncbi.nlm.nih.gov/33894014/
- Reference implementation (MIT license, cited in paper Section 4.6): https://github.com/Tanaka-Group/AD_QSP_model

Miyano 2022 develops a quantitative systems pharmacology (QSP) model of atopic dermatitis (AD) pathogenesis that couples 14 ODE states — skin barrier integrity, infiltrated pathogens, four helper T-cell subsets (Th1, Th2, Th17, Th22), seven cytokines (IL-4, IL-13, IL-17A, IL-22, IL-31, IFN-gamma, TSLP), and OX40L — with drug-effect modules for nine biologics: dupilumab (anti-IL-4Ra, blocks IL-4 and IL-13 signalling), lebrikizumab and tralokinumab (both anti-IL-13), secukinumab (anti-IL-17A), fezakinumab (anti-IL-22), nemolizumab (anti-IL-31 receptor), tezepelumab (anti-TSLP), GBR 830 (anti-OX40L), and recombinant IFN-gamma (rIFNg). The efficacy endpoint is the Eczema Area and Severity Index (EASI = 36 * (pathogens + 1 - barrier), an algebraic function of the barrier-integrity and pathogen states, bounded in the 0-72 EASI range).

The model layer for each biologic is a fixed inhibition fraction on the effective cytokine concentration seen by downstream regulatory targets (r_inhibit = 0.99 for antibodies, r_inhibit = 0.44 for tralokinumab from an e_a2 = 0.4396 multiplier that reflects tralokinumab's dual-receptor mechanism, or an additive c_rifng = 210 for recombinant IFN-gamma). Miyano 2022 model-based meta-analysis of nine placebo-controlled trials tuned 51 log-normally distributed patient-specific parameters (mu, sigma) that drive between-patient heterogeneity in pathophysiologic rate constants.

## Population

The QSP layer describes system-level AD pathogenesis independently of individual demographics. The MBMA layer pools nine randomised placebo-controlled trials of investigational biologics for moderate-to-severe AD (Miyano 2022 Table 1; 663 placebo + 848 drug-arm patients across 9 trials). Virtual patients are defined only by their draws from the 51 log-normal parameter distributions; no age / weight / race / sex covariates are used in the model.

The same information is available programmatically via `readModelDb("Miyano_2022_atopicDermatitis_qsp")()$population`.

## Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()` entry in `inst/modeldb/therapeuticArea/Miyano_2022_atopicDermatitis_qsp.R`. The table below collects the structural equations and the drug-effect encoding in one place.

| Element | Source |
|---|---|
| 14 ODEs (skin barrier, pathogens, Th1-Th22, 7 cytokines, OX40L) | Paper Sec 2.2 (14 ODEs, 51 parameters); reference implementation AD_QSP_model.py lines 155-174 |
| EASI algebraic observation (36*(pathogens + 1 - barrier)) | AD_QSP_model.py EASI() function lines 61-65 |
| Drug-effect on cytokines ((1 - r_inhibit) * c_actual) | Paper Sec 2.3 (r_inhibit = 0.99 for antibodies) |
| Tralokinumab e_a2 = 0.4396 multiplier | Paper Sec 4.2 (Discussion, "e_a2 = 0.44"); AD_QSP_model.py line 140 |
| rIFNg additive skin concentration (c_rifng = 210) | Paper Sec 2.3 (Eq. for c_IFNg); AD_QSP_model.py line 230 |
| Placebo / trial-arm activation of k3 skin-barrier recovery | Paper Sec 3.3 ("k3 corresponds to placebo effects"); AD_QSP_model.py line 91 |
| 51 mu values (log-scale means, rows 1-51) | Reference implementation mu.csv (author-supplied, MIT-licensed) |
| 51 sigma values (log-scale SDs, rows 1-51) | Reference implementation sigma.csv (author-supplied, MIT-licensed) |
| 11 elimination-rate mu values | Paper Sec 2.4 states these were derived from in-vivo human serum half-lives listed in Table S5. Table S5 lives in the paper's Wiley supplement, which is not currently on disk; the numerical values here are taken from mu.csv (author-supplied companion) so no substitution from external data is used. |
| Table S2 baseline biological-factor levels (paper seed IC's for the steady-state run) | Paper Sec 2.4 references Table S2; the numerical seed values (0.5931, 0.4069, 3.1, …) are reproduced from AD_QSP_model.py lines 183-198. |
| 9 drug arms (dose regimens, patient counts) | Paper Table 1 |
| Figure 3 (baseline levels, EASI trajectories, EASI-75) | Paper Sec 3.2, Figure 3 |

## Parameter terminology: mu, sigma, and eta (a naming bridge)

Before reaching for software vocabulary, it helps to anchor on a general, software-agnostic distinction between two *levels* of noise that any population model separates: **patient-variability-level noise** and **measurement-level noise**. Patient-variability-level noise (also called between-subject variability, or BSV) is variability in the parameters that govern each individual's own dynamics — two patients running the exact same ODE structure but with different rate constants. Measurement-level noise (residual/observation error) is variability between what the model predicts for one individual and what is actually observed for that same individual at that same time point — assay noise, timing imprecision, or unmodeled short-term fluctuation. This QSP model only has the first kind: the 51 log-normal parameter draws (`mu`, `sigma`) create between-patient variability, and from there each patient's ODEs are solved deterministically, with no measurement-noise layer added on top of the prediction.

With that concept in place, the software vocabulary below is one continuous lineage, not two competing conventions: NONMEM's classic `$THETA`/`$OMEGA`/`$ETA`/`$SIGMA` control-stream records defined this naming decades ago for population PK/PD, and nlmixr2's R-native `ini()`/`model()` DSL (compiled to an ODE solver by rxode2) deliberately reuses those same four words so PMx modelers feel at home — it isn't a rival naming scheme. The paper's own `mu.csv`/`sigma.csv` files, however, predate and are independent of that PMx vocabulary, and happen to reuse the word "sigma" for something that maps onto a *different* one of the four NONMEM/nlmixr2 terms. That's the collision worth untangling:

- **QSP Terminology: mu/sigma** (Miyano 2022 reference implementation, `mu.csv`/`sigma.csv`): for each of the 51 rate constants, `mu` is the log-scale population mean and `sigma` is the log-scale standard deviation describing between-patient (patient-variability-level) variability in that rate constant.
- **PMx Terminology: THETA/OMEGA/ETA** (nlmixr2's `ini()` block, compiled by rxode2): `mu` becomes a log-scale THETA (e.g. `lk1 <- 0.537`), and `sigma` is folded into the variance of an ETA (e.g. `etalk1 ~ 0.628849`, where `0.628849 = 0.793^2`) — i.e. the paper's sigma lands in NONMEM's OMEGA slot, not its SIGMA slot, despite the name overlap. Each patient's rate constant is then reconstructed as `k1 <- exp(lk1 + etalk1)` in `model()`.
- **PMx Terminology: SIGMA** (residual/observation error variance): a different concept from either of the above — measurement or assay noise added on top of an individual prediction when fitting observed data. It's absent from this model entirely, since this QSP model has no measurement/observation-error layer; it only ever reports the deterministic EASI trajectory computed from the ODE states. This is why rxode2 prints `Warning message: No sigma parameters in the model` when you load it: that warning refers to this inherited NONMEM SIGMA slot being empty, not to the paper's `sigma.csv` (which, again, nlmixr2 compiles into OMEGA/ETA instead).

The table below anchors on the four canonical PMx building blocks (THETA, OMEGA, ETA, and residual-error SIGMA) as the primary key, since these are the exact terms you'll meet again when sampling a population PK model — only the paper's QSP-specific naming changes underneath them.

| PMx Term | This model's notation | QSP Term (paper) | Role |
|---|---|---|---|
| Individual parameter | `k1 <- exp(lk1 + etalk1)` | the patient-specific rate constant | What actually enters `d/dt(...)` for that patient |
| THETA (fixed effect) | `lk1` | `mu` (mu.csv, log scale) | Population-typical value of the rate constant |
| ETA (random effect) | `etalk1` | not named separately in the paper — implicit in "patients sampled from mu, sigma" | One virtual patient's deviation, drawn ~ N(0, OMEGA) |
| OMEGA (variance) | `0.628849` in `etalk1 ~ 0.628849` | `sigma^2` (sigma.csv, squared) | Between-patient variability of that rate constant |
| SIGMA (residual error) | absent — no `sigma()` block here | n/a — no measurement/observation layer in this QSP model | Measurement-level noise (absent here); do not confuse with the paper's `sigma.csv`, which is patient-variability-level noise mapped to OMEGA above, not this |

`zeroRe()` sets every ETA to exactly 0, so all 51 rate constants collapse to their THETA (population `mu`) values — this is the "typical patient" used in the steady-state check below. The later cohort-simulation section keeps the ETAs active (a fresh random draw per virtual patient, `nSub = n_per_arm`) to generate patient-to-patient heterogeneity.

## Visualizing THETA, ETA, and between-patient variability

The THETA/OMEGA/ETA machinery in the table above becomes concrete once it's sampled. Two figures below use the model's first parameter (`k1`, skin barrier recovery rate) and two model objects created here: `mod` (full between-patient variability) and `mod_typ` (the same model with `zeroRe()`, i.e. all etas fixed at 0 — the "typical patient" used throughout the rest of this document).

```{r load-model}
mod <- readModelDb("Miyano_2022_atopicDermatitis_qsp")
mod_typ <- mod |> rxode2::zeroRe()
```

The first figure samples 300 virtual patients' `etalk1` draws directly (without solving any ODEs) and shows both the raw ETA (log scale, centred on 0) and the resulting individual `k1` (natural scale, log-normal, centred on the THETA-derived typical value):

```{r theta-eta-distribution, fig.width=8, fig.height=4}
set.seed(123)
n_demo <- 300
lk1 <- 0.537
omega_k1 <- 0.628849  # = 0.793^2; sigma.csv row 1 squared -> etalk1 ~ 0.628849 in ini()

eta_k1 <- rnorm(n_demo, mean = 0, sd = sqrt(omega_k1))
k1_indiv <- exp(lk1 + eta_k1)
k1_typical <- exp(lk1)
cat("Typical-patient k1 =", k1_typical, "\n")

df_theta_eta <- rbind(
  data.frame(quantity = "ETA (etalk1, log scale)", value = eta_k1),
  data.frame(quantity = "individual k1 = exp(lk1 + etalk1)", value = k1_indiv)
)

ggplot(df_theta_eta, aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7, colour = "white") +
  facet_wrap(~ quantity, scales = "free", nrow = 1) +
  labs(x = NULL, y = "Count",
       title = "THETA + ETA -> individual parameter (300 simulated virtual patients)",
       subtitle = paste0("ETA ~ N(0, OMEGA = ", round(omega_k1, 3), "); typical-patient k1 = exp(lk1) = ", round(k1_typical, 3)))
```

The same random ETAs that produced this parameter-level spread propagate through all 14 ODEs to produce different simulated trajectories per virtual patient, even though every patient shares identical THETAs. The second figure simulates 12 virtual patients (fresh ETA draws, `mod`) alongside the typical patient (`mod_typ`, all ETAs = 0):

```{r theta-eta-spaghetti, fig.width=7, fig.height=4.5}
set.seed(123)
n_spaghetti <- 12
ev_demo <- rxode2::et(seq(0, 100, by = 2)) |> rxode2::et(id = seq_len(n_spaghetti))
ev_demo$cmt <- "barrier"
ev_demo_typ <- rxode2::et(seq(0, 100, by = 2)) |> rxode2::et(id = 1L)
ev_demo_typ$cmt <- "barrier"

sim_demo <- rxode2::rxSolve(mod, events = ev_demo, nSub = n_spaghetti) |> as.data.frame()
sim_demo_typ <- rxode2::rxSolve(mod_typ, events = ev_demo_typ) |> as.data.frame()

ggplot(sim_demo, aes(x = time, y = EASI, group = id)) +
  geom_line(colour = "grey60", alpha = 0.6) +
  geom_line(data = sim_demo_typ, aes(x = time, y = EASI), inherit.aes = FALSE,
            colour = "black", linewidth = 1.1) +
  labs(x = "Time (weeks)", y = "EASI",
       title = "Same THETAs, different ETAs: 12 virtual patients vs. the typical patient",
       subtitle = "Grey = individual virtual patients (random eta draws); black = typical patient (zeroRe(), eta = 0)",
       caption = "This same eta-sampling mechanism generates between-subject variability in a population PK model.")
```

The grey trajectories fan out from (and back toward) the black typical-patient curve purely because of ETA sampling — nothing about the THETAs or the structural ODEs changes between patients. This THETA-fixed / ETA-random pattern is exactly what you will see again when sampling from a population PK model: THETA sets the population-typical PK parameter (e.g. CL, V), OMEGA sets its between-subject variance, and ETA is each simulated subject's individual draw.

## Steady-state check (typical patient)

The paper simulates 1000 weeks of drug-free evolution to establish the model's steady state, then applies drug at t = 0 of the trial. The vignette replicates that steady-state run for the typical patient (`zeroRe()`, i.e. all etas set to zero, using the `mod_typ` object created above) and confirms the state settles into a stable attractor. The initial conditions written into the model are Table-S2-informed seed values (not the model's fitted steady state), so the check is that the state DOES drift over the equilibration run and then stops drifting.

```{r steady-state}
# Steady-state run: no drug, 1000 weeks, weekly observations. cmt = "barrier"
# anchors the observation timing to a real ODE state (see the model's
# paper_specific_compartments list); rxode2 returns EASI (the algebraic
# observable) as a column regardless.
ev_ss <- rxode2::et(seq(0, 1000, by = 4)) |>
  rxode2::et(id = 1L)
ev_ss$cmt <- "barrier"

ss <- rxode2::rxSolve(mod_typ, events = ev_ss) |> as.data.frame()
tail(ss[, c("time", "barrier", "pathogens", "il13", "th2", "EASI")], 5)
```

At t = 1000 weeks the typical-patient steady-state EASI should sit in the moderate-to-severe range consistent with the modelled AD population (Miyano 2022 Figure 3A baselines).

```{r steady-state-plot}
ss |>
  dplyr::filter(time <= 200) |>
  ggplot(aes(x = time)) +
  geom_line(aes(y = EASI, colour = "EASI")) +
  geom_line(aes(y = 20 * barrier, colour = "20*barrier")) +
  geom_line(aes(y = pathogens, colour = "pathogens")) +
  scale_colour_manual(values = c("EASI" = "black", "20*barrier" = "steelblue", "pathogens" = "firebrick")) +
  labs(x = "Time (weeks)", y = "State value",
       colour = NULL,
       title = "Typical-patient equilibration to steady state",
       caption = "Drift-and-settle from Table-S2 seed IC's to the model's fitted attractor.")
```

## How drug effect is currently modelled (no PK compartment)

Before reproducing Figure 3B, it is worth being explicit about what "drug" means inside this QSP model, because it is a much simpler mechanism than a population-PK model provides.

There is no PK compartment, dose, or absorption/distribution/elimination process for any of the nine biologics. Each drug arm is instead a **constant, patient-independent inhibition fraction** applied for the entire 24-week trial window, starting at t = 0. Concretely (reference implementation `AD_QSP_model.py`, drug-effect block, lines ~143-148):

```python
IL4  = (1 - de[1]) * c[6]
IL13 = (1 - de[2] * ea2) * c[7]   # ea2 = 0.4396 for tralokinumab, else 1
IL17 = (1 - de[3]) * c[8]
IL22 = (1 - de[4]) * c[9]
IL31 = (1 - de[5]) * c[10]
TSLP = (1 - de[6]) * c[12]
OX40 = (1 - de[7]) * c[13]
IFNg = c[11] + de[8]              # additive, not multiplicative
```

In the `ini()`/`model()` DSL used throughout this vignette, `de[1]` through `de[7]` are the `r_il4`, `r_il13`, ..., `r_ox40l` parameters (0.99 for full antibody blockade, 0 for no effect), and `de[8]` is `c_rifng` (additive rIFNg skin concentration). Three consequences follow directly from this design:

- **Step-function onset.** The inhibition fraction is either "on" (0.99) or "off" (0) — there is no ramp-up while a drug distributes and reaches its target, because there is no distribution to model. A virtual patient's cytokine inhibition is exactly as strong in week 1 as it is in week 24.
- **No between-patient PK variability.** `r_il4`, `r_il13`, etc. are `fixed()` THETAs in `ini()` — every patient in an arm receives the identical value. All of the patient-to-patient heterogeneity you see in the spaghetti plot above comes entirely from the 51 QSP-mechanism ETAs (skin barrier turnover, Th-cell differentiation, cytokine production, ...), never from variability in how much drug reaches its target.
- **Not the same thing as the cytokine elimination-rate parameters.** The model does contain 11 real elimination-rate constants (`ld1`, `ld8`, `ld9`-`ld17`; Table S5-sourced half-lives) — but these describe how fast the *cytokines themselves* (IL-4, IL-13, ...) are cleared from the physiological system once produced, not how fast the *drug* is cleared from the body. There is no analogous drug-clearance parameter anywhere in this model.

This is a defensible simplification for reproducing steady-state efficacy endpoints (mean EASI, EASI-75 at 24 weeks) from a model-based meta-analysis, since it collapses drug PK and target engagement into one tuned scalar per drug. It does mean the model cannot answer PK-linked questions -- dose/schedule optimization, missed-dose sensitivity, or how much of the response lags behind absorption -- the way a real population-PK/PD model would.

The dedicated PK section at the end of this vignette ("Illustrative Dupilumab population PK") builds a real THETA/OMEGA/ETA-based one-compartment PopPK submodel for dupilumab, links its simulated concentration to a time-varying version of `r_il4`/`r_il13`, and compares the resulting EASI-75 trajectory against this fixed-inhibition mechanism for the same virtual cohort.

## Reproduce Figure 3B: mean EASI trajectories for the 9 drugs

Each drug corresponds to a distinct setting of the drug-scenario parameters exposed in `ini()`. One events dataframe is built per drug arm (typical patient, weekly observations, no dosing events because the model treats drug exposure as a constant inhibition fraction applied for the whole treatment period).

```{r drug-params}
# Drug-scenario parameter table. Each row = one arm. r_placebo = 1 for every
# clinical arm (placebo effect activates the k3 skin-barrier recovery term);
# r_placebo = 0 was the steady-state run above.
drug_params <- tibble::tribble(
  ~drug,          ~r_placebo, ~r_il4, ~r_il13, ~r_il17, ~r_il22, ~r_il31, ~r_tslp, ~r_ox40l, ~c_rifng, ~e_a2,
  "Placebo",       1,         0,       0,      0,      0,      0,       0,       0,        0,        1,
  "Dupilumab",     1,         0.99,   0.99,    0,      0,      0,       0,       0,        0,        1,
  "Lebrikizumab",  1,         0,      0.99,    0,      0,      0,       0,       0,        0,        1,
  "Tralokinumab",  1,         0,      0.99,    0,      0,      0,       0,       0,        0,        0.4396,
  "Secukinumab",   1,         0,       0,     0.99,    0,      0,       0,       0,        0,        1,
  "Fezakinumab",   1,         0,       0,      0,     0.99,    0,       0,       0,        0,        1,
  "Nemolizumab",   1,         0,       0,      0,      0,     0.99,     0,       0,        0,        1,
  "Tezepelumab",   1,         0,       0,      0,      0,      0,      0.99,     0,        0,        1,
  "GBR830",        1,         0,       0,      0,      0,      0,       0,      0.99,      0,        1,
  "rIFNg",         1,         0,       0,      0,      0,      0,       0,       0,        210,      1
)
```

The equilibrated typical-patient steady state (from the previous chunk) is used as the initial condition for every drug arm (i.e. the model settles for 1000 weeks with no drug, then starts the trial). Simulating for 24 weeks matches the paper's efficacy-endpoint window.

```{r drug-arms}
# Take the last row of the ss run as the equilibrated baseline. It has the
# 14 state columns.
state_cols <- c("barrier", "pathogens", "th1", "th2", "th17", "th22",
                "il4", "il13", "il17", "il22", "il31", "ifng", "tslp", "ox40l")
baseline_typ <- ss[nrow(ss), state_cols, drop = FALSE]

typical_by_drug <- function(row) {
  ev <- rxode2::et(seq(0, 24, by = 0.5)) |> rxode2::et(id = 1L)
  ev$cmt <- "barrier"
  p <- as.numeric(row[c("r_placebo", "r_il4", "r_il13", "r_il17", "r_il22",
                        "r_il31", "r_tslp", "r_ox40l", "c_rifng", "e_a2")])
  names(p) <- c("r_placebo", "r_il4", "r_il13", "r_il17", "r_il22",
                "r_il31", "r_tslp", "r_ox40l", "c_rifng", "e_a2")
  sim <- rxode2::rxSolve(mod_typ, events = ev, params = p,
                         inits = unlist(baseline_typ)) |>
    as.data.frame()
  sim$drug <- row[["drug"]]
  sim
}

typ_all <- purrr::map_dfr(seq_len(nrow(drug_params)), function(i)
  typical_by_drug(drug_params[i, ]))

# Compute %improved EASI relative to each drug's own baseline (t = 0 within
# the trial). The paper's Figure 3B shows %improved EASI over time.
typ_all <- typ_all |>
  group_by(drug) |>
  mutate(pct_improved_EASI = 100 * (first(EASI) - EASI) / first(EASI)) |>
  ungroup()
```

```{r figure3B, fig.width=8, fig.height=5}
ggplot(typ_all, aes(x = time, y = pct_improved_EASI, colour = drug)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time on treatment (weeks)",
       y = "%improved EASI (typical patient)",
       colour = NULL,
       title = "Typical-patient EASI trajectory by drug arm",
       caption = "Replicates the qualitative pattern of Miyano 2022 Figure 3B (means).")
```

The relative ordering (dupilumab and lebrikizumab -> highest efficacy; secukinumab and rIFNg -> smallest EASI improvement) reproduces the Figure 3B pattern qualitatively; the exact numerical trajectories in the paper are population means with IIV, not typical-patient projections.

## Small virtual-patient cohort for EASI-75

For a stochastic cohort simulation, a virtual population (50 patients) is sampled and a single long trajectory per patient per drug arm is simulated: the first 1000 weeks equilibrate without drug (r_placebo = 0), then the drug switches on at week 1000. The 24-week drug arm response is computed relative to the per-patient EASI at t = 1000 (each patient's own equilibrated baseline). Cohort cap: 200 per arm; the vignette uses 50 per arm x 10 arms = 500 patient-arms.

Rather than time-varying drug covariates, 2 sims are run per patient per drug: an initial no-drug 1000-week equilibration (shared across all drugs), then a 24-week drug arm from the shared typical baseline. Per-patient parameter draws (from the etas) drive per-patient trajectory heterogeneity during the drug arm; using the typical steady-state IC as the shared starting point is a small simplification vs the paper's per-patient equilibrated IC's.

```{r cohort-sim}
n_per_arm <- 50L
mod_stoch <- mod  # keeps eta variability

# Use the typical-patient equilibrated state as the shared IC for all
# virtual patients in every drug arm. This differs from the paper (which
# equilibrates each virtual patient individually) but keeps runtime linear
# in cohort size and captures the drug-response heterogeneity of interest.
baseline_typ_named <- as.numeric(baseline_typ[1, ])
names(baseline_typ_named) <- names(baseline_typ)

sim_arm <- function(drug_row) {
  ev <- rxode2::et(seq(0, 24, by = 1)) |>
    rxode2::et(id = seq_len(n_per_arm))
  ev$cmt <- "barrier"
  p <- as.numeric(drug_row[c("r_placebo", "r_il4", "r_il13", "r_il17",
                             "r_il22", "r_il31", "r_tslp", "r_ox40l",
                             "c_rifng", "e_a2")])
  names(p) <- c("r_placebo", "r_il4", "r_il13", "r_il17", "r_il22",
                "r_il31", "r_tslp", "r_ox40l", "c_rifng", "e_a2")
  sim <- rxode2::rxSolve(mod_stoch, events = ev,
                         nSub = n_per_arm,
                         params = p,
                         inits = baseline_typ_named) |>
    as.data.frame()
  sim$drug <- drug_row[["drug"]]
  sim
}

cohort_all <- purrr::map_dfr(seq_len(nrow(drug_params)), function(i)
  sim_arm(drug_params[i, ]))

# %improved EASI relative to each patient's t = 0 (in-arm) baseline.
baseline_easi <- cohort_all |>
  dplyr::filter(time == 0) |>
  dplyr::select(drug, id, EASI_baseline = EASI)
cohort_all <- cohort_all |>
  dplyr::left_join(baseline_easi, by = c("drug", "id")) |>
  dplyr::mutate(pct_improved = 100 * (EASI_baseline - EASI) / EASI_baseline)
```

```{r figure3C, fig.width=8, fig.height=5}
# EASI-75: fraction of patients per arm per timepoint achieving >75% EASI improvement.
easi75 <- cohort_all |>
  dplyr::group_by(drug, time) |>
  dplyr::summarise(EASI75 = 100 * mean(pct_improved > 75, na.rm = TRUE),
                   .groups = "drop")

ggplot(easi75, aes(x = time, y = EASI75, colour = drug)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time on treatment (weeks)", y = "EASI-75 responder rate (%)",
       colour = NULL,
       title = "EASI-75 by drug arm (50 virtual patients per arm)",
       caption = "Replicates Miyano 2022 Figure 3C at reduced cohort size.")
```

At week 24, the ordering matches the paper's Figure 3C (dupilumab and lebrikizumab -> highest EASI-75 responder rate; secukinumab and rIFNg near the bottom). Exact numerical agreement requires the paper's 1000-per-arm cohort and 1000 repeated simulated trials (Figure 3 caption); the vignette's 50-per-arm point estimate is illustrative.

## Assumptions and deviations

- Paper supplement (SI Section 3 with the S1 Table of ODEs, Table S3 with r_inhibit derivations, Tables S4/S5 with the 51 mu and sigma parameter values) is not currently on disk. The parameter values (mu.csv, sigma.csv) and the ODE equations used here were reproduced from the authors' MIT-licensed reference implementation at https://github.com/Tanaka-Group/AD_QSP_model, which the paper explicitly cites as the canonical implementation (Section 4.6: "The code of the QSP model is available at …"). The failed auto-acquisition of the Wiley supplement is recorded in `needs_acquisition.jsonl` for future manual pickup; when the supplement lands, the two sources should be cross-checked for numerical agreement.
- Between-patient variability interpretation. The paper describes each patient as a draw from a log-normal distribution on each of the 51 rate constants (mu, sigma). This is encoded here as 51 independent `etal<name> ~ sigma^2` diagonal random effects. No off-diagonal covariance is reported in the paper or the reference implementation. The paper's MBMA scope is between-patient variability in pathophysiologic parameters, not between-study heterogeneity in the population-PK sense.
- Drug-effect parameters exposed as `ini()` overrides. The nine drug arms are simulated by setting `r_il4`, `r_il13`, …, `c_rifng`, `e_a2` at `rxSolve()` time via `params = c(...)`. Default values reproduce the drug-free steady state (all r_* = 0; r_placebo = 0; e_a2 = 1; c_rifng = 0). Users can compose novel therapies by setting arbitrary combinations of these fractions (Miyano's Figure 6 explores this space to identify simultaneous IL-13 + IL-22 inhibition as the best-performing bispecific candidate for dupilumab poor responders).
- Vignette cohort size. The vignette uses n = 50 per arm to stay inside the pkgdown render time budget. The paper's Figure 3 uses 1000 virtual patients per arm and 1000 repeated simulated trials for 95% CIs. The qualitative shape (drug ordering, response-rate direction) reproduces at n = 50; the exact numerical Figure-3 CIs do not.
- Vignette skips PKNCA. The paper's efficacy endpoint (EASI-75 at 24 weeks) is not an NCA quantity; there is no dose-and-eliminate PK profile to integrate. Following the endogenous / mechanistic validation pattern, the vignette uses steady-state check and figure replication instead.
- Table S3 r_inhibit derivations from IC50 and skin drug concentrations are not on disk. The r_inhibit = 0.99 for antibodies and r_inhibit = 0.44 for tralokinumab (via e_a2 = 0.4396) are used as stated in the paper's Section 2.3 and Section 4.2 respectively; the derivations of these two values from IC50 + skin concentration data (paper Section 3.4 and Table S3) cannot be independently verified from the on-disk sources.
- Table S5 elimination-rate half-lives are not on disk. The paper states these were derived from published in-vivo human serum half-lives. The numerical mu values used for `ld1`, `ld8`, `ld9`, `ld10-d17` come from `mu.csv` in the reference implementation; the underlying half-life citations from Table S5 are not visible for cross-reference.

## Illustrative Dupilumab population PK: linking THETA/OMEGA/ETA across PK and QSP PD

The section above ("How drug effect is currently modelled") described why this QSP model, as published, uses a constant inhibition fraction. As a self-contained illustration of what a *linked* PK/QSP-PD model looks like, this section builds a small one-compartment population-PK (PopPK) submodel for dupilumab, simulates a large virtual cohort's dupilumab concentration-time profiles, converts each patient's own concentration to a time-varying `r_il4`/`r_il13` inhibition fraction (an Emax link), and re-simulates EASI-75 for a larger cohort than used above, so a fixed-inhibition vs. PK-driven comparison is apples-to-apples.

**Provenance note.** The dosing regimen below (600 mg SC loading dose, then 300 mg SC every 2 weeks) is the FDA-labeled adult dupilumab regimen for atopic dermatitis and is treated as a citable fact. The PK model's THETA values (`ka`, `CL`, `V`) and OMEGA (between-patient variability) magnitudes are, by contrast, **illustrative only** -- chosen to produce a physiologically plausible mAb-like absorption/elimination shape (few-day SC absorption half-life, few-week elimination half-life), not fit to any specific published dupilumab population-PK dataset. Treat the resulting concentration values as arbitrary relative units, not mg/L.

```{r pk-dupi-model}
# Illustrative one-compartment, first-order-absorption PopPK model for
# dupilumab. Time unit = weeks (matches the rest of this vignette).
# THETA/OMEGA magnitudes are illustrative mAb-like values, NOT fit to any
# specific published dataset (see provenance note above).
pk_dupi_fn <- function() {
  ini({
    lka <- log(1.73)   # THETA: ka ~= 1.73/week -> absorption t1/2 ~= 0.4 week (~2.8 d), typical SC-mAb range
    lcl <- log(0.35)   # THETA: CL ~= 0.35 (V-units)/week -> elimination t1/2 ~= 2 weeks with lv below
    lv  <- log(1.5)    # THETA: V, arbitrary relative volume units (not literature-derived)
    etaka ~ 0.16        # OMEGA: ka IIV, ~40% CV -- illustrative SC-mAb absorption variability
    etacl ~ 0.09        # OMEGA: CL IIV, ~30% CV -- illustrative mAb clearance variability
    etav  ~ 0.04        # OMEGA: V IIV, ~20% CV -- illustrative mAb volume variability
  })
  model({
    ka <- exp(lka + etaka)
    cl <- exp(lcl + etacl)
    v  <- exp(lv + etav)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <-  ka * depot - (cl / v) * central
    conc <- central / v
  })
}
pk_dupi <- rxode2::rxode2(pk_dupi_fn)
pk_dupi
```

This mirrors exactly the THETA/OMEGA/ETA pattern used throughout this vignette for the QSP-mechanism parameters above: `lka`/`lcl`/`lv` are population-typical log-scale THETAs, `etaka`/`etacl`/`etav` are the between-patient random effects, and each simulated patient's own `ka`, `cl`, `v` triple is `exp(THETA + ETA)`, exactly as `k1 <- exp(lk1 + etalk1)` was for the QSP skin-barrier recovery rate.

```{r pk-dupi-cohort}
n_cohort <- 200L               # 4x the n_per_arm = 50 used in the Figure 3C cohort above
t_grid <- seq(0, 24, by = 0.5) # finer than the 1-week grid above, to resolve the early PK-driven onset ramp

set.seed(2026)
# Labeled adult AD regimen: 600 mg SC loading dose at t=0, then 300 mg SC
# every 2 weeks (ii=2, addl=11 covers the 24-week window).
ev_pk <- rxode2::et(amt = 600, time = 0, cmt = "depot", id = seq_len(n_cohort)) |>
  rxode2::et(amt = 300, time = 2, ii = 2, addl = 11, cmt = "depot") |>
  rxode2::et(t_grid)

pk_sim <- rxode2::rxSolve(pk_dupi, events = ev_pk, nSub = n_cohort) |> as.data.frame()
summary(pk_sim$conc)
```

Across the `r n_cohort`-patient cohort, simulated concentration ranges from `r round(min(pk_sim$conc), 1)` to `r round(max(pk_sim$conc), 1)` (arbitrary relative units) over the 24-week window, rising sharply during the first dosing interval and then oscillating around an accumulated multiple-dose plateau.

### Simulated PK profile (median + prediction interval "VPC-style" band)

A visual predictive check (VPC) normally overlays observed-data percentiles on top of a simulated prediction interval to check that a fitted population-PK model reproduces the observed concentration distribution. No observed dupilumab concentration dataset is attached to this repository, so the plot below is a **simulation-only** population time-course: at each simulated time point across the `r n_cohort`-patient cohort, it summarises the median and the 5th-95th percentile interval of simulated concentration, in the same spirit as a VPC's simulated-percentile band, but with nothing observed to compare it against. Treat it as a check that the illustrative THETA/OMEGA magnitudes above produce a plausible population profile shape (fast multi-dose accumulation, then a plateau), not as a fit diagnostic.

```{r pk-dupi-vpc, fig.width=8, fig.height=5}
pk_vpc_summary <- pk_sim |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    p05    = quantile(conc, 0.05, na.rm = TRUE),
    median = quantile(conc, 0.50, na.rm = TRUE),
    p95    = quantile(conc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(pk_vpc_summary, aes(x = time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), colour = "steelblue4", linewidth = 1) +
  labs(x = "Time (weeks)", y = "Simulated dupilumab concentration (arbitrary relative units)",
       title = paste0("Illustrative dupilumab population PK: simulated concentration-time profile (n = ", n_cohort, ")"),
       subtitle = "Median (line) and 5th-95th percentile interval (ribbon) across the simulated virtual cohort",
       caption = "Simulation-only band -- no observed concentration data overlaid (illustrative THETA/OMEGA, see provenance note above); not a fit-validating VPC.")
```

The ribbon widens visibly around each dosing event before narrowing again as the cohort's absorption-phase concentrations converge toward the next trough, which is the expected between-patient spread from the `etaka`/`etacl`/`etav` OMEGAs layered on top of the deterministic multi-dose accumulation pattern.

Concentration is converted to a time-varying inhibition fraction with a simple Emax model:

$$r\_il4(t) = r\_il13(t) = E_{max} \cdot \frac{C(t)}{EC_{50} + C(t)}$$

```{r pk-emax-link}
Emax <- 0.99  # matches this model's existing antibody-inhibition ceiling (r_il4 = r_il13 = 0.99 above)
EC50 <- 25    # illustrative; arbitrary relative units matching `conc` above; not fit to target-engagement data

pk_cov <- pk_sim |>
  dplyr::transmute(id, time,
                   r_il4  = Emax * conc / (EC50 + conc),
                   r_il13 = Emax * conc / (EC50 + conc))
```

`Emax = 0.99` is chosen to exactly match this model's existing antibody-inhibition ceiling (the same 0.99 used for the fixed-inhibition Dupilumab arm above), so any difference between the two mechanisms reflects only the time-course, not a different maximum effect. `EC50 = 25` (same arbitrary relative units as `conc`) is an illustrative choice -- not fit to any target-engagement dataset -- picked from this cohort's simulated concentration range so that the inhibition fraction visibly ramps up over the first 1-2 weeks of absorption before plateauing near the 0.99 ceiling, which is the qualitative onset-delay behaviour a real PK/PD link is expected to show.

```{r pk-driven-cohort}
# QSP observation-only event table (same time grid + cohort size as the PK
# simulation above), merged with each patient's own PK-derived time-varying
# r_il4/r_il13 covariate. A data column sharing an ini() parameter's name
# overrides that parameter at each row, and rxode2 uses it as a time-varying
# covariate instead of the population-constant default.
ev_pk_driven <- rxode2::et(t_grid) |>
  rxode2::et(id = seq_len(n_cohort))
ev_pk_driven$cmt <- "barrier"
ev_pk_driven <- as.data.frame(ev_pk_driven) |>
  dplyr::left_join(pk_cov, by = c("id", "time"))

# Remaining drug-scenario parameters held at their Dupilumab constant values
# (only r_il4/r_il13 are time-varying here).
p_pk_other <- c(r_placebo = 1, r_il17 = 0, r_il22 = 0, r_il31 = 0,
                r_tslp = 0, r_ox40l = 0, c_rifng = 0, e_a2 = 1)

# Same seed as the fixed-inhibition solve below -> identical 200 QSP-parameter
# draws for both mechanisms, isolating drug-effect mechanism as the only
# difference between the two cohorts.
set.seed(2028)
pk_driven_sim <- rxode2::rxSolve(mod_stoch, events = ev_pk_driven,
                                 params = p_pk_other,
                                 inits = baseline_typ_named) |>
  as.data.frame()
```

```{r pk-fixed-cohort}
# Same cohort size + time grid, current fixed-inhibition mechanism, as the
# direct comparator (this is the same Dupilumab row already in drug_params
# above, just re-run at n_cohort = 200 instead of n_per_arm = 50).
ev_fixed <- rxode2::et(t_grid) |> rxode2::et(id = seq_len(n_cohort))
ev_fixed$cmt <- "barrier"
p_fixed <- c(r_placebo = 1, r_il4 = 0.99, r_il13 = 0.99, r_il17 = 0, r_il22 = 0,
             r_il31 = 0, r_tslp = 0, r_ox40l = 0, c_rifng = 0, e_a2 = 1)

set.seed(2028)
fixed_sim <- rxode2::rxSolve(mod_stoch, events = ev_fixed,
                             params = p_fixed,
                             inits = baseline_typ_named) |>
  as.data.frame()
```

```{r pk-easi75-compare, fig.width=8, fig.height=5}
easi75_from <- function(sim, label) {
  baseline_easi <- sim |> dplyr::filter(time == 0) |> dplyr::select(id, EASI_baseline = EASI)
  sim |>
    dplyr::left_join(baseline_easi, by = "id") |>
    dplyr::mutate(pct_improved = 100 * (EASI_baseline - EASI) / EASI_baseline) |>
    dplyr::group_by(time) |>
    dplyr::summarise(EASI75 = 100 * mean(pct_improved > 75, na.rm = TRUE), .groups = "drop") |>
    dplyr::mutate(mechanism = label)
}

easi75_compare <- dplyr::bind_rows(
  easi75_from(pk_driven_sim, "PK-driven (time-varying Emax)"),
  easi75_from(fixed_sim, "Fixed inhibition (current model)")
)

ggplot(easi75_compare, aes(x = time, y = EASI75, colour = mechanism)) +
  geom_line(linewidth = 0.9) +
  labs(x = "Time on treatment (weeks)", y = "EASI-75 responder rate (%)",
       colour = NULL,
       title = paste0("Dupilumab EASI-75: fixed inhibition vs. PK-driven mechanism (n = ", n_cohort, " per mechanism)"),
       caption = "Same 200 QSP-parameter draws for both mechanisms; only the drug-effect mechanism differs.")
```

```{r pk-easi75-summary, include=FALSE}
get_val <- function(df, wk, mech) df$EASI75[df$time == wk & df$mechanism == mech]
wk2_pk     <- round(get_val(easi75_compare, 2, "PK-driven (time-varying Emax)"), 0)
wk2_fixed  <- round(get_val(easi75_compare, 2, "Fixed inhibition (current model)"), 0)
wk24_pk    <- round(get_val(easi75_compare, 24, "PK-driven (time-varying Emax)"), 0)
wk24_fixed <- round(get_val(easi75_compare, 24, "Fixed inhibition (current model)"), 0)
```

At week 2, the PK-driven mechanism's EASI-75 responder rate (`r wk2_pk`%) lags behind the fixed-inhibition mechanism (`r wk2_fixed`%), because the PK-driven inhibition fraction is still ramping up through the SC absorption phase while the fixed-inhibition mechanism is (unrealistically) already at full 0.99 blockade from t = 0. By week 24 the two mechanisms have not fully converged (`r wk24_pk`% vs. `r wk24_fixed`%): the accumulated multi-dose concentration plateaus a little below full saturation of the Emax relationship, so the PK-driven mechanism's effective ceiling (~0.93, per the concentration range above) stays slightly under the fixed mechanism's flat 0.99. The qualitative lesson generalizes beyond dupilumab: a model that treats drug effect as an instantaneous constant will systematically overstate early-treatment response relative to a model that links a real absorption/distribution/elimination profile to target engagement -- the discrepancy is largest during the first 1-2 half-lives of drug exposure, and any residual gap at steady state reflects how close the chosen `EC50` lets concentration get to saturating the Emax relationship.

Assumptions specific to this illustrative PK section:

- All PK THETA/OMEGA values (`lka`, `lcl`, `lv` and their etas) are illustrative mAb-like magnitudes, not fitted to a specific published dupilumab population-PK dataset -- only the labeled dosing regimen (600 mg SC load, 300 mg SC q2w) is a citable fact.
- `conc` is reported in arbitrary relative units (`central`/`V` with an arbitrary `V`), not mg/L; `EC50 = 25` is chosen in those same units purely to produce a visible ~1-2 week onset ramp, not fit to a target-engagement/receptor-occupancy dataset.
- The PK ETAs (`etaka`, `etacl`, `etav`) and the 51 QSP-mechanism ETAs are drawn independently per virtual patient (the same `id` links one patient's PK draw to their own QSP draw, but no correlation between the two is assumed or estimated) -- a simplification, since no published joint PK/QSP-severity distribution exists for this model.
- Only `r_il4` and `r_il13` are made time-varying here (dupilumab's two cytokine targets); the remaining six drug-effect parameters are held at their Dupilumab constant values (0 for all, per `drug_params` above).

Developed by Richard Hooijmaijers, Matthew Fidler, Bill Denney, Kiranmayi Vedantham.

---

Adapted by Blerta Shtylla and Luke Fostvedt from the original rxode2 tutorial by Richard Hooijmaijers, Matthew Fidler, Bill Denney, and Kiranmayi Vedantham (cited above).
